Who Gets Cared For and Why? Fertility Shocks, Intra-Household Investment Reallocation, and the Quantity–Quality Trade-Off in the Philippines

### 2025-08-07 docs/paper.Rmd
revise paper title, abstract, intro, and literature review
Updated the paper's title, short title, acknowledgements, and abstract to reflect a new focus on fertility shocks and intra-household investment in the Philippines. Substantially revised the introduction and literature review for clarity, conciseness, and to better align with the new research questions. Added new LaTeX header includes and made minor technical corrections throughout the RMarkdown file. Added a new preamble.tex file and updated references.

---
title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households. I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No. 003—which effectively banned the provision of modern contraceptives in all city-run health facilities. Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy. The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

Empirical investigations of the quantity-quality trade-off have focused primarily on educational outcomes. Most studies examine how increases in maternal fertility—or equivalently, the number of siblings a child has—affect school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other dimensions of child quality–such as health at birth and early childhood health investments—are early-onset or biologically constrained, and they tend to be less amenable to remediation later in life. For instance, babies born with very low birth weight face higher risks of infant mortality and long-term developmental deficits that cannot be fully reversed \citep{cook2015understanding}. Similarly, a child who misses critical early vaccinations is more vulnerable to infectious diseases that can kill or disable them, which results to lifelong negative health impacts \citep{costa2024child}. Such early-life health setbacks may permanently impair cognitive development and physical growth. This limits long-run productivity regardless of subsequent improvements in schooling or incom

@@

I use this natural experiment to estimate the causal impact of increased family size on early-childhood health outcomes. The analysis uses nationally representative data from multiple rounds of the Philippine Demographic and Health Survey (DHS), which include detailed information on household structure, maternal characteristics, and child health indicators. The primary outcomes are birth weight (measured in grams, when reported) and the mother’s subjective assessment of the child’s size at birth—both of which serve as indicators of neonatal health status. In addition, I assess whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment. These outcomes are widely recognized indicators of child well-being and reflect both biological and behavioral dimensions of parental investment in early development. The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the absence of the contraceptive ban, trends in child health outcomes in Manila would have evolved in parallel with those in the comparison cities. Second, no other policy changes or economic shocks in Manila during the study period coincided with the timing and scope of the contraceptive policy. I probe these assumptions using falsification tests, placebo comparisons, and robustness checks that include city-specific time trends, mother fixed effects, and controls for baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, perceived birth size, and vaccination uptake). Finally, I examine heterogeneity in these effects across subgroups defined by maternal education, household wealth, and access to prenatal care. These characteristics proxy for a household’s resources and its capacity to buffer the impact of increased family size on child health.

@@

In what follows, I review the theoretical foundations of the Q-Q model and its evolution in the literature. I begin with the static models of Becker \citep{becker1960economic} and Becker–Lewis \citep{becker1973interaction}, which first formalized the trade-off within a household utility maximization framework. We then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and unified growth models, notably Galor and Weil \citep{galor2000population} and Galor and Moav \citep{galor2002natural}, which integrate the Q–Q mechanism into a general theory of demographic and economic transformation.

@@

A further refinement of the quantity-quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions. The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \). In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines. According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health. These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.

@@

\[
\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
\]

with a corresponding condition for \( q \). Here, \( \alpha \in [0,1] \) represents the wife’s bargaining weight. When \( \alpha \) increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.

@@

The theoretical question is how these dimensions interact with the quantity decision. If parents allocate a budget across, say, schooling \( q_{\text{edu}} \) and nutrition/health \( q_{\text{health}} \) for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them). In some models, health and education are complementary: a healthier child benefits more from education, and an educated mother might raise a healthier child. This complementarity can amplify the Q–Q trade-off—investing in one dimension (health) increases the returns to investing in the other (education), so a high-quality strategy becomes more focused on fewer children.

@@

Beyond survival, early-life health indicators such as birth weight, perceived size at birth, and vaccination status are important dimensions of child quality, especially in low-income settings. When resources are limited, parents may struggle to meet the basic needs of each child. Empirical evidence aligns with the quantity–quality framework: children from larger families often have worse health outcomes at birth and are less likely to receive core immunizations. For example, recent multi-country data show that the share of completely unimmunized children—so-called “zero-dose” children—increases from 10.5% among those with no siblings to 17.2% among those with four or more \citep{costa2024child}. Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth. These patterns suggest that as fertility increases, the health of each child may suffer due to more limited parental investment. Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth

Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia, where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs. Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:

\[
H = g(F, M)
\]

and \( H \) enters either utility directly or the child’s future productivity. Parents then choose \( n \), \( F \), \( M \), and possibly schooling per child. Larger \( n \) reduces \( F \) and \( M \) per child, lowering \( H \); if parents place high value on \( H \), or if \( H \) enhances returns to education, they will opt for smaller families.

@@

The results are striking. The contraceptives ban led to a significant rise in births and family size in Manila city relative to the control areas. Correspondingly, children born in Manila during the ban era experienced a sizable decline in educational attainment – clear evidence of a Q–Q trade-off precipitated by the shock. In the authors’ words, the policy-driven increase in family size “provide[s] evidence of a quality–quantity trade-off”: larger families, forced by the ban, resulted in lower schooling outcomes per child. By leveraging an actual anti-contraception policy, this study offers causal confirmation that constraints on family planning can worsen child quality. Together with Orbeta’s findings, it reinforces the relevance of Becker’s hypothesis in the Philippine setting: whether via unintended fertility (Manila’s case) or ordinary variation in sibling sex mix, more children mean fewer resources per child and thus worse education outcomes. Consistent with theory, Filipino parents facing exogenous increases in quantity were unable to maintain the same level of quality per child.

@@

This study builds upon the foundational framework of the quantity-quality (Q–Q) trade-off, first articulated in \citet{becker1960economic} and later refined in \citet{becker1973interaction}. In this model, fertility and child quality are jointly determined under a full-income constraint, such that increases in the number of children—without a corresponding expansion in household resources—necessitate a reduction in the average investment per child. While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, perceived size at birth, and uptake of essential vaccinations.

@@

To formalize this relationship, consider a household with total income \( I \), choosing nutritional investments \(\kappa_j\) for each of \( n \) children, along with consumption \( Z \) unrelated to child quality:

\[
I = \sum_{j=1}^{n} \kappa_j + Z.
\]

In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers. If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight or perceived neonatal size, and children may be less likely to receive complete immunization schedules.

The first hypothesis derived from this framework is that an exogenous increase in fertility leads to a decline in per-child health investment, particularly in low-income urban households with little financial slack. A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight, smaller perceived size at birth, and reduced uptake of key childhood vaccinations. These hypotheses describe a causal pathway through which fertility shocks may contribute to early deprivation and potentially reinforce intergenerational cycles of poor health and limited human capital.

@@

The primary data source for this study is the Philippine Demographic and Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds. These nationally representative surveys provide detailed child- and household-level information, including maternal characteristics, fertility behavior, and early-childhood health outcomes. Importantly, they are the only Philippine DHS waves with available GPS cluster data, which I use to approximate geographic location and assign respondents to treatment and control groups. Because city-level identifiers are not released in the public-use files, I rely on coarse geographic assignment based on cluster coordinates. Clusters located within or near the administrative boundaries of the City of Manila are classified as “treated,” while clusters located in other parts of the National Capital Region (NCR) serve as the control group. In robustness checks, I restrict the control group to Quezon City or expand it to the rest of NCR excluding both Manila and Quezon City.

The analytic sample includes children under five years of age living with their mothers at the time of the survey. I focus on three early-life health outcomes that reflect both biological endowments and parental investment: (1) birth weight, reported in grams when available; (2) the mother’s subjective assessment of birth size; and (3) vaccination uptake, defined as receipt of key early-childhood immunizations, including BCG, DPT, polio, and measles. These outcomes are consistently available across all three waves and serve as proxies for child quality in the context of the quantity–quality trade-off.

The 1998 DHS is used as the primary pre-policy baseline wave because it captures children born before the implementation of Manila’s contraceptive ban in 2000. The 2008 DHS serves as the post-policy round to capture outcomes for children born well after the ban took effect. I also include the 2003 DHS in the main specification, recognizing that children in this round may have been born shortly before or after the policy’s introduction. I treat 2003 as a transition-period dataset and conduct sensitivity checks to assess the robustness of findings to its inclusion. I exclude the 1993 DHS from the analysis due to the absence of GPS data, which makes it impossible to reliably assign treatment status.

To support the identification strategy, I use DHS summary indicators on contraceptive use disaggregated by region and city. These data confirm a sharp and sustained decline in modern contraceptive use in Manila following the implementation of Executive Order No. 003, relative to surrounding cities, validating the policy as a plausibly exogenous fertility shock. I also considered using data from the Philippine National Nutrition Survey (NNS), which includes rich anthropometric and dietary data. However, the NNS microdata are only available from 2003 onward, which limits their value for pre-policy analysis. As such, the DHS remains the most suitable dataset for this study.

\subsection{Identification Strategy}

To estimate the causal effect of increased fertility on early-childhood health outcomes, I exploit a policy shock generated by Executive Order No. 003, issued in 2000 by the mayor of Manila. This order restricted access to modern contraceptives in all city-run health facilities and remained in effect for nearly a decade. The policy was implemented only in the City of Manila and did not affect neighboring municipalities within the National Capital Region (NCR), which continued to provide modern family planning services. This spatially localized and time-bound restriction created a natural experiment that allows for a quasi-experimental research design. I define the treatment group as children residing in GPS clusters located within or near the Manila boundary, and the control group as children in other NCR municipalities. Given the limitations of the DHS GPS data, which are displaced for confidentiality and do not include exact city identifiers, I use a coarser geographic classification to approximate exposure. Robustness checks vary the control group to include Quezon City only, or the rest of NCR excluding both Manila and Quezon City.

The core of my empirical strategy is a difference-in-differences (DiD) design that compares changes in child outcomes in Manila before and after the ban relative to changes in similar urban areas that were unaffected. The key assumption is that, in the absence of the policy, trends in early-childhood health outcomes would have followed similar trajectories across treatment and control areas. This approach accounts for both fixed differences between cities and nationwide time trends.

To address potential violations of the parallel trends assumption—such as Manila-specific shocks that may have affected all children regardless of family size—I implement a triple-differences strategy that leverages variation in exposure by maternal age. Because fertility declines sharply with age, older women were biologically less likely to respond to the contraceptive ban. For example, mothers aged 45–49 at the time of the policy were near the end of their reproductive years and thus unlikely to experience a fertility increase. By comparing changes in child outcomes across maternal age cohorts within Manila and control cities, I can account for time-varying, location-specific confounders that affect all families, while isolating the policy's effect on family size and its consequences.

I further restrict my main analysis to children born before the policy was implemented. Although these children were not conceived due to the ban, they may still have been indirectly affected by the fertility increase through resource dilution within households. This strategy avoids selection bias related to unobservable differences in the endowments of “marginal children” born because of the policy. As argued by Rosenzweig and Zhang (2009), when parents reinforce rather than compensate for birth endowments, estimates based on older siblings provide a lower bound of the true effect of fertility on child quality. Empirical evidence from the Philippines supports this pattern of reinforcement, as shown by Baguet and Dumas (2015). Taken together, these strategies allow me to estimate the causal impact of fertility shocks on early-life health while addressing key concerns about internal validity and sample composition.

\subsection{Sample Selection}

The analytic sample is drawn from the Philippine Demographic and Health Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003, and 2008. I restrict the sample to children under age five (0–59 months) with non-missing information on key outcomes: birth weight (in grams, when available), perceived size at birth, and vaccination status. Because the Manila contraceptive ban was enacted in 2000 and likely implemented unevenly at first, I define children born in 2001 and later as post-policy and those born before 1997 as pre-policy. Children born between 1997–2000 are excluded in the main specification but retained in robustness checks.

I include all live births regardless of birth order or plurality to preserve variation in sibship size. Twins and higher-order multiple births are flagged for sensitivity checks but not excluded. Each child is linked to their biological mother using maternal identifiers, allowing construction of sibship size, birth order, and maternal age. Exposure to the policy is defined geographically: children are considered treated if they were born in the City of Manila, and untreated if born in other municipalities within the National Capital Region (NCR). Because GPS data are coarse in some years, geographic exposure is defined at the city level rather than using precise coordinates. Households outside NCR are excluded from the main analysis to ensure urban comparability and avoid confounding by rural fertility and health dynamics.

@@

The primary outcome variables reflect early-life health and investment. These include: (1) birth weight in grams, as recorded by the mother when available; (2) the mother’s subjective assessment of birth size (categorized as very small, small, average, large, or very large); and (3) indicators for whether the child received core early-life vaccinations, including BCG, DPT1–3, polio doses, and measles. A binary variable indicating whether the child received all recommended vaccines is also constructed as a summary measure of postnatal health investment.

The main explanatory variable is sibship size, defined as the total number of surviving children ever born to the mother at the time of the child’s birth, as reported in the birth history module. This variable captures the resource dilution mechanism central to the quantity–quality trade-off. To assess whether the contraceptive ban caused an increase in sibship size, I also construct an indicator for policy exposure, coded as 1 if the child was born in the City of Manila and in or after 2001 (i.e., post-ban), and 0 otherwise. In extended specifications, I interact this treatment indicator with maternal age at birth to capture heterogeneity in fertility effects across age cohorts.

>>>>

---
title: "Who Gets Cared For and Why? Fertility Shocks, Intra-Household Investment Reallocation, and the Quantity–Quality Trade-Off in the Philippines"
short: "Spillovers from Policy-Induced Fertility"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible through the support of the Schupf Fellowship at Amherst College. I am deeply thankful to Dr. Axel Schupf ’57 for his generosity and vision in establishing this program, which annually supports a cohort of twenty rising sophomores and juniors in independent summer research. I feel truly privileged to have been selected through this competitive process and to have spent eight weeks immersed in a project that both challenged and inspired me. I am especially grateful to my mentor, Prof. Caroline Theoharides, for her mentorship throughout this project—and more broadly, for sparking my interest in economic research and its potential to serve the public good in the Philippines. I also wish to thank Faculty Director Professor Amelie Hastie and the many campus partners whose institutional support made this experience possible. Finally, I am grateful to the Departments of Economics and Mathematics & Statistics for equipping me with the analytical and conceptual tools that made this research possible—and for preparing me to take on independent research with confidence and curiosity. All views and any errors are my own.
abstract: |
  This paper investigates how increases in fertility, driven by reduced access to contraception, affect parental investment in child health. I use Executive Order No. 003—a 2000 policy in Manila that restricted access to modern contraceptives—as an exogenous fertility shock. Drawing on pooled data from the 1998, 2003, and 2008 Philippine Demographic and Health Surveys, I estimate two-stage least squares models that use policy exposure to instrument for sibship size. Child health inputs fall into two categories: prenatal (proxied by birth weight) and postnatal (measured by the total number of vaccine doses received by age five). The first-stage results indicate that affected mothers had, on average, 0.20 more living children at the time of a birth. In the second stage, each additional sibling is associated with a reduction of 22.97 vaccine doses received by the index child, a decline large enough to prevent completion of the recommended immunization schedule. The estimated effect on birth weight is −8,676 grams, though statistically insignificant Reduced-form estimates are directionally consistent: policy exposure is associated with a 0.91-dose drop in vaccine uptake and a 349.60-gram decrease in birth weight. While instrumental variable estimates are limited by a weak first stage, the magnitude and sign of the coefficients support the hypothesis that larger sibship sizes reduce health investments per child. The results suggest that postnatal investments, particularly those requiring sustained parental effort, are more sensitive to fertility shocks than prenatal outcomes.
output: 
  rticles::aea_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
bibliography: references.bib
header-includes:
  - \usepackage{pdfpages}
  - \usepackage{float}
  - \usepackage{makecell}
  - \setlength{\parskip}{0pt}       % Remove vertical space between paragraphs
  - \setlength{\parindent}{15pt}    % Indent first line of each paragraph
---

@@

Empirical investigations of the quantity-quality trade-off have focused primarily on educational outcomes. Most studies examine how increases in maternal fertility—or equivalently, the number of siblings a child has—affect school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other dimensions of child quality–such as health at birth and early childhood health investments—are early-onset or biologically constrained, and they tend to be less amenable to remediation later in life. For instance, babies born with very low birth weight face higher risks of infant mortality and long-term developmental deficits that cannot be fully reversed \citep{cook2015understanding}. Similarly, a child who misses critical early vaccinations is more vulnerable to infectious diseases that can kill or disable them, which results to lifelong negative health impacts \citep{costa2024child}. Such early-life health setbacks may permanently impair cognitive development and physical growth. This limits long-run productivity regardless of subsequent improvements in schooling or income.

@@

I use this natural experiment to estimate the causal impact of increased family size on early-childhood health outcomes. The analysis uses nationally representative data from multiple rounds of the Philippine Demographic and Health Survey (DHS), which include detailed information on household structure, maternal characteristics, and child health indicators. The primary outcomes are birth weight (measured in grams, when reported) and whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment. These outcomes are widely recognized indicators of child well-being and reflect both biological and behavioral dimensions of parental investment in early development. The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the absence of the contraceptive ban, trends in child health outcomes in Manila would have evolved in parallel with those in the comparison cities. Second, no other policy changes or economic shocks in Manila during the study period coincided with the timing and scope of the contraceptive policy. I probe these assumptions using falsification tests, placebo comparisons, and robustness checks that include city-specific time trends, mother fixed effects, and controls for baseline demographic differences. The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, and vaccination uptake). Finally, I examine heterogeneity in these effects across subgroups defined by maternal education, household wealth, and access to prenatal care. These characteristics proxy for a household’s resources and its capacity to buffer the impact of increased family size on child health.

@@

In what follows, I review the theoretical foundations of the Q-Q model and its evolution in the literature. I begin with the static models of Becker \citep{becker1960economic} and Becker–Lewis \citep{becker1973interaction}, which first formalized the trade-off within a household utility maximization framework. I then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and unified growth models, notably Galor and Weil \citep{galor2000population} and Galor and Moav \citep{galor2002natural}, which integrate the Q–Q mechanism into a general theory of demographic and economic transformation.

@@

A further refinement of the quantity-quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions. The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility $n$ and child quality $q$. In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines. According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health. These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.

@@

$$
\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
$$

with a corresponding condition for $q$. Here, $\alpha \in [0,1]$ represents the wife’s bargaining weight. When $\alpha$ increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.

@@

The theoretical question is how these dimensions interact with the quantity decision. If parents allocate a budget across, say, schooling $q_{\text{edu}}$ and nutrition/health $q_{\text{health}}$ for each child, then having more children forces cutbacks in both dimensions (unless parents reallocate across them). In some models, health and education are complementary: a healthier child benefits more from education, and an educated mother might raise a healthier child. This complementarity can amplify the Q–Q trade-off—investing in one dimension (health) increases the returns to investing in the other (education), so a high-quality strategy becomes more focused on fewer children.

@@

Beyond survival, early-life health indicators such as birth weight and vaccination status are important dimensions of child quality, especially in low-income settings. When resources are limited, parents may struggle to meet the basic needs of each child. Empirical evidence aligns with the quantity–quality framework: children from larger families often have worse health outcomes at birth and are less likely to receive core immunizations. For example, recent multi-country data show that the share of completely unimmunized children—so-called “zero-dose” children—increases from 10.5% among those with no siblings to 17.2% among those with four or more \citep{costa2024child}. Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight. These patterns suggest that as fertility increases, the health of each child may suffer due to more limited parental investment. Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight.

Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia, where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs. Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health $H$ depends on food $F$ and medical care $M$, such that:

$$
H = g(F, M)
$$

and $H$ enters either utility directly or the child’s future productivity. Parents then choose $n$, $F$, $M$, and possibly schooling per child. Larger $n$ reduces $F$ and $M$ per child, lowering $H$; if parents place high value on $H$, or if $H$ enhances returns to education, they will opt for smaller families.

@@

The results are striking. The contraceptives ban led to a significant rise in births and family size in Manila city relative to the control areas. Correspondingly, children born in Manila during the ban era experienced a sizable decline in educational attainment – clear evidence of a Q–Q trade-off precipitated by the shock. In the authors’ words, the policy-driven increase in family size “provide[s] evidence of a quality–quantity trade-off”: larger families, forced by the ban, resulted in lower schooling outcomes per child.

@@

This study builds upon the foundational framework of the quantity-quality (Q–Q) trade-off, first articulated in \citet{becker1960economic} and later refined in \citet{becker1973interaction}. In this model, fertility and child quality are jointly determined under a full-income constraint, such that increases in the number of children—without a corresponding expansion in household resources—necessitate a reduction in the average investment per child. While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, and uptake of essential vaccinations.

@@

To formalize this relationship, consider a household with total income $I$, choosing nutritional investments $\kappa_j$ for each of $n$ children, along with consumption $Z$ unrelated to child quality:

$$
I = \sum_{j=1}^{n} \kappa_j + Z.
$$

In the presence of a fertility shock, holding $I$ constant, the average $\kappa_j$ must fall unless offset by changes in parental behavior or external transfers. If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight, and children may be less likely to receive complete immunization schedules.

The first hypothesis derived from this framework is that an exogenous increase in fertility leads to a decline in per-child health investment, particularly in low-income urban households with little financial slack. A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight and reduced uptake of key childhood vaccinations. These hypotheses describe a causal pathway through which fertility shocks may contribute to early deprivation and potentially reinforce intergenerational cycles of poor health and limited human capital.

@@

```{r}
#| label: load-libraries
#| include: false

# Core Data Import and Manipulation 
library(tidyverse)     # Includes ggplot2, dplyr, readr, etc.
library(data.table)    # For fast data handling
library(haven)         # For reading Stata (.dta) files
library(janitor)       # For cleaning variable names and tables

# Variable Labels and Factors 
library(labelled)      # For working with labelled survey variables
library(forcats)       # For factor manipulation

# Table Rendering and Summarization 
library(gtsummary)     # For descriptive summary tables
library(gt)            # For rendering gtsummary tables
library(kableExtra)    # For LaTeX/HTML table styling
library(knitr)         # For knitting reports and inline tables
library(broom)         # For tidying model output

# Modeling and Estimation 
library(fixest)        # For fixed effects and IV regressions
library(modelsummary)  # For regression output formatting
library(AER)           # For ivreg() — standard 2SLS instrumental variable estimation
library(MASS)          # For polr()
library(ordinal)       # For clm() – flexible ordered logit

# Plotting and Utilities 
library(purrr)         # For functional programming
library(scales)        # For axis and color formatting in ggplot2
library(extrafont)     # For font support in graphics

```

```{r}
#| label: data-loading
#| include: false

# Load cleaned DHS child-level data with geographic identifiers ---
dhs1 <- readRDS("../data-analysis/clean-data/dhs_kr_children_with_geo.rds")

# Prepare cleaned variables and identifiers ---
dhs2 <- dhs1 |>
  filter(survey_year %in% c(1998, 2003, 2008)) |>
  haven::zap_labels() |>
  mutate(
    # Standardize birth year to 4-digit format
    birth_year = ifelse(birth_year < 100, 1900 + birth_year, birth_year),

    # Clean and recode birth weight and perceived birth size
    birth_weight_grams = na_if(birth_weight, birth_weight %in% c(9996, 9998, 9999)),
    birth_size_clean = na_if(birth_size, birth_size %in% c(8, 9)),

    birth_size_factor = case_when(
      birth_size == 1 ~ "Very large",
      birth_size == 2 ~ "Larger than average",
      birth_size == 3 ~ "Average",
      birth_size == 4 ~ "Smaller than average",
      birth_size == 5 ~ "Very small",
      TRUE ~ NA_character_
    ) |> factor(levels = c(
      "Very small", "Smaller than average", "Average",
      "Larger than average", "Very large"
    )),

    # Flag NCR either via region code or admin name
    ncr = region_dhs == 14 | tolower(admin_region_name) == "national capital region",

    # Manually fix municipality names for known Manila GPS coordinates (2008 only)
    municipality_name = case_when(
      survey_year == 2008 &
        latitude %in% c(14.6055, 14.56937, 14.56748, 14.56274, 14.59441) &
        longitude %in% c(120.9485, 120.96649, 120.97882, 120.98432, 121.01443) ~ "Manila",
      TRUE ~ municipality_name
    ),

    # Identify Manila municipalities
    manila = grepl("manila", tolower(municipality_name)),

    # Define additional flags (after correcting birth_year)
    other_ncr = ncr & !manila,
    post_policy = birth_year >= 2000,
    manila_post = manila & post_policy
  )

# Filter for NCR cities only
dhs2_ncr <- dhs2 |> filter(ncr)
```

This study draws on two primary data sources: the Philippine Demographic and Health Surveys (DHS) and the Integrated Public Use Microdata Series (IPUMS) census data. The DHS provides detailed child- and household-level information on fertility behavior, maternal characteristics, and early childhood health outcomes. I use three rounds of the DHS—1998, 2003, and 2008—which are the only Philippine DHS waves with publicly available GPS cluster data. These data are used to classify respondents into treatment and control groups. Because city-level identifiers are not included in the public-use DHS files, I rely on approximate geographic assignment using cluster coordinates. Clusters located within or near the administrative boundaries of the City of Manila are classified as treated, while clusters located elsewhere in the National Capital Region (NCR) serve as the control group.

The DHS analytic sample is limited to children under five years of age who were living with their mothers at the time of the survey. I examine three core early-life health outcomes: (1) reported birth weight (in grams); (2) maternal perception of birth size; and (3) vaccination uptake, measured as the receipt of key early childhood immunizations (BCG, DPT, polio, and measles). These outcomes are consistently available across all three survey rounds and are commonly used as proxies for child quality in studies of the quantity–quality trade-off.

To complement the DHS analysis and support the identification strategy, I construct a panel of individual-level microdata from the IPUMS harmonized Philippine censuses (1990, 2000, and 2010). These nationally representative surveys provide a snapshot of women of reproductive age (15–49) and contain detailed fertility histories—including total children ever born, births in the past year, and number of surviving children. I limit the sample to women ages 15–49 and construct a composite fertility index combining several measures: total children ever born, births in the past year, number of children under age five, and the age of the youngest child. By comparing Manila with other provinces and municipalities in NCR, I test whether long-run fertility patterns reflect the contraceptive policy shock.

The 1998 DHS serves as the primary pre-policy baseline, capturing births prior to the issuance of Executive Order No. 003. The 2008 DHS is the main post-policy round, while the 2003 DHS is treated as a transition period. I exclude the 1993 DHS due to the absence of geospatial identifiers. In contrast, the census years 1990 and 2000 bracket the policy’s onset, while the 2010 census provides post-policy fertility outcomes. This combination of sources enables both short-run and long-run assessments of the policy’s impact on fertility and child investment.

To validate the policy shock, I use DHS region- and city-level contraceptive use indicators, which confirm a sharp and sustained decline in modern contraceptive use in Manila relative to other NCR cities. These patterns reinforce the plausibility of Executive Order No. 003 as an exogenous shock to fertility preferences and access. I also considered the Philippine National Nutrition Survey (NNS) as an auxiliary source, but its limited pre-2003 coverage restricts its utility for identifying pre-policy trends. For this reason, the combined DHS and IPUMS datasets offer the most comprehensive and temporally balanced data available for the study's objectives.

@@

The primary outcome variables reflect early-life health and investment. These include: (1) birth weight in grams, as recorded by the mother when available; and (2) indicators for whether the child received core early-life vaccinations, including BCG, DPT1–3, polio doses, and measles.

The main explanatory variable is sibship size, defined as the total number of surviving children ever born to the mother at the time of the child’s birth, as reported in the birth history module. This variable captures the resource dilution mechanism central to the quantity–quality trade-off. To isolate exogenous variation in sibship size, I construct an instrument for fertility based on policy exposure: an interaction of a Manila indicator and a post-ban cohort indicator. Specifically, children born in the City of Manila in or after 2000 are coded as treated. In extended specifications, I interact this treatment indicator with maternal age at birth to allow for heterogeneity in fertility responses across age cohorts, considering the fact that older women likely exhibited lower responsiveness due to reduced fecundity.
